---
title: "Final Project - Machine Learning"
author: "KristinButler"
date: "2/2/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Background

  Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

# Summary
The training data was first cleaned up where columns with no data and identifiers were removed from the data.  Cleaning up the data resulted in reduction from 160 to 53 variables.  The data was divided into training and test data.  Next the only the variables that showed high correlation (>0.8) were selected or 11 variables.  From this data frame a, four models were tested and the model with the highest accuracy selected for final prediction.  The random forest model showed highest accuracy or 97% accuracy. 


# Load packages
```{r, warning = F, comment = F, cache= F}

library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(randomForest)
library(corrplot)
library(gbm)
library(rpart.plot)
library(e1071)
library(tidyr)
library(dplyr)
```


# Getting and Cleaning data

## Import data 
```{r}
dat <- read.csv("C:/Users/Kristin.Butler/Desktop/Coursera/MachineLearn/Final/pml-training.csv", header = T, na.strings = c("NA", "")) 

pre <- read.csv("C:/Users/Kristin.Butler/Desktop/Coursera/MachineLearn/Final/pml-testing.csv", header = T, na.strings = c("NA", ""))

dim(dat)
dim(pre)
head(dat)

```
The training data has 160 columns with 19622 rows.  This is allot of variable and need to be reduced so that analysis don't take up to much time.  Additionally, variable with NA's and identifiers need to be removed before the analysis.  


## Clean data 
Remove columns with no data, and identifiers.  

```{r}
#Remove columns with no data in it
dat1 <- dat[, colSums(is.na(dat)) == 0]
dim(dat1)


pre1 <- pre[, colSums(is.na(pre)) == 0]
dim(pre1)

head(dat1)

#Remove identifiers from data set
dat2 <- dat1[,-c(1:7)]
pre2 <- pre1[,-c(1:7)]
dim(dat2)
dim(pre2)
```
Removing no data columns and identifiers reduced the variables to 53. 


# Divide into training and test data

The training data was devided into training set and testing set.  
```{r}
inTrain = createDataPartition(dat2$classe, p = 3/4)[[1]]

training = dat2[ inTrain,]

testing = dat2[-inTrain,]

dim(training)
dim(testing)
```



## Reduce variables

The data was reduced by selecting the varibales that had high or >0.8 correlation.  This was done by creating a correlation matrix and selecting the variables with >0.8 correlation.  As a results the varibales were reduced from 53 to 14.  A data frame with only these 14 variables was created for model testing.  


```{r}
#Correlation
library(kernlab)
M <- abs(round(cor(training[, -53]),2))
head(M, 6)

#Make a correlation plot
par(mar=c(5, 2, 1, 2))
corrplot(M, type = "upper", order = "hclust", tl.col = "black", tl.srt = 45, mar=c(1,1,0,1), tl.cex = 0.50)

#Find highly correlated variable to include in model
highlyCorrelated = findCorrelation(M, cutoff=0.80)

nam <- names(training)[highlyCorrelated]
nam2 <- c("classe", nam)


#Select columns with keywords in main correlations
training2 <- dplyr::select(training,nam2)
dim(training2)


testing2 <- select(testing,nam2)
dim(testing2)

```

## Modeling

Four models were created (random forest, oosting and 2x regression trees), and their accuracy compared.  

### Random Forest Modeling
```{r}
set.seed(2334)
# Random forest
mod1 <- train(classe ~., data = training2, method = "rf", prox = T)
P1<- predict(mod1, testing2)
c1 <- confusionMatrix(P1, testing2$classe)$overall['Accuracy']


```

### Predicting with trees
```{r}
set.seed(2334)
mod2 <- train(classe ~., method ="rpart", data = training2)
P2<- predict(mod2, testing2)
c2 <- confusionMatrix(P2, testing2$classe)$overall['Accuracy']

library(rattle)
fancyRpartPlot(mod2$finalModel)

```

## Boosting
```{r, warning = F,comment = F, cache= F}
mod3 <- train(classe ~., method ="gbm", data = training2)
P3<- predict(mod3, testing2)
c3 <- confusionMatrix(P3, testing2$classe)$overall['Accuracy']

```


```{r}
mod4 <- train(classe ~., method ="lda", data = training2)
P4<- predict(mod4, testing2)
c4 <- confusionMatrix(P4, testing2$classe)$overall['Accuracy']

```

###  Model comparison
```{r}
Correlation <- c(c1, c2, c3, c4)

model <- c("rf", "rpart", "gbm", "lda")

comb2 <- as.data.frame(Correlation)
comb3 <- as.data.frame(comb2, model)
round(comb3, 2)
```
Comparing all four models, the random forest was the most accurate one and was used for prediction on the test data. 




## Final model
The random forest model was used for prediction and below are the results in table from.  

```{r}
pre1 <- dplyr::select(pre, nam)
predict <- predict(mod1, pre1)
table(predict)
as.matrix(predict)
```
